---
title: "CodeBook"
author: "Mike Holmquest"
date: "8/31/2020"
output:
  html_document: default
  pdf_document: default
---
<div id="instructions">
Project Introduction
</div>

One of the most exciting areas in all of data science right now is wearable computing - see for example this article . Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

Here are the data for the project:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip


<div id="instructions">
Review criteria
</div>

1. The submitted data set is tidy.
2. The Github repo contains the required scripts.
3. GitHub contains a code book that modifies and updates the available codebooks with the data to indicate all the variables and summaries calculated, along with units, and any other relevant information.
4. The README that explains the analysis files is clear and understandable.
5. The work submitted for this project is the work of the student who submitted it.

<div id="instructions">
Getting and Cleaning Data Course Project
</div>

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. You will be graded by your peers on a series of yes/no questions related to the project. You will be required to submit: 
1) a tidy data set as described below, 
2) a link to a Github repository with your script for performing the analysis, and 
3) a code book that describes the variables, the data, and any transformations or work that you performed to clean up the data called CodeBook.md. 
You should also include a README.md in the repo with your scripts. This repo explains how all of the scripts work and how they are connected.


You should create one R script called run_analysis.R that does the following.

Merges the training and the test sets to create one data set.
Extracts only the measurements on the mean and standard deviation for each measurement.
Uses descriptive activity names to name the activities in the data set
Appropriately labels the data set with descriptive variable names.
From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
Good luck!

## Load data program

```{r, echo=FALSE}

library(dplyr)
```

The files now need to be downloaded from the internet from a zip file and loaded into the files folder.

##Download Zip File

```{r, echo = TRUE}

if (!file.exists("UCIDataset.zip")){
        fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
        download.file(fileURL, destfile = "UCIDataset.zip" , method="curl")
}  

if (!file.exists("UCIDataset")) { 
        unzip("UCIDataset.zip") 
}
```

Once all the files are downloaded they need to be loaded into the environment to use in R Studio

## Load all data frames from project files

```{r, echo = TRUE}

features <- read.table("UCI HAR Dataset/features.txt", col.names = c("n","functions"))
activities <- read.table("UCI HAR Dataset/activity_labels.txt", col.names = c("code", "activity"))
subject_test <- read.table("UCI HAR Dataset/test/subject_test.txt", col.names = "subject")
x_test <- read.table("UCI HAR Dataset/test/X_test.txt", col.names = features$functions)
y_test <- read.table("UCI HAR Dataset/test/y_test.txt", col.names = "code")
subject_train <- read.table("UCI HAR Dataset/train/subject_train.txt", col.names = "subject")
x_train <- read.table("UCI HAR Dataset/train/X_train.txt", col.names = features$functions)
y_train <- read.table("UCI HAR Dataset/train/y_train.txt", col.names = "code")
```

The first task is to merge the data sets together. To do this you either need to column bind first and then row bind, or row bind first, and then column bind. I chose to row bind first. and then create a new file named "Merged_UCI_Data" that had all of the needed raw dat in it. 

## Merge all training and test data sets

```{r, echo = TRUE}


X <- rbind(x_train, x_test)
Y <- rbind(y_train, y_test)
Subject <- rbind(subject_train, subject_test)
Merge_UCI_Data <- cbind(Subject, Y, X)

```

Once the raw data is merged you need to then extract the identification data (subject and activity) and also any variables that have mean, or standard deviation data. 

## Extract only the mean and standard deviation measurement


```{r, echo=TRUE}
TidyextractData <- Merge_UCI_Data %>%
        select(subject, code, contains("mean"), contains("std"))
```

The next task was to change the code variable to more accurately describe the activity that it represents. by beginning to "tidy" the data with accurate labels it makes it easier to understand and analyze. 
## Use descriptive activity names in the dataset

```{r, echo=TRUE}
## Use descriptive activity names in the dataset
TidyextractData$code <- activities[TidyextractData$code, 2]

## Write the Extracted Tidy Data to a file
write.table(TidyextractData, "TidyExtractData.txt", row.name = FALSE)
```

Now that we have the data within the columns tidy, our next task is to tidy the labels for the columns. I decided to first loook at the current names of the columns to find out what I could change. I found a lot of abbreviations were being used so I tried to relabel what I could to clarify the variables. 

## Preview current column names

```{r}
names(TidyextractData)
```

## Appropriately label the data set with descriptive names

```{r, echo=TRUE}
names(TidyextractData)[1] = "SubjectID"
names(TidyextractData)[2] = "ActivityID"
names(TidyextractData) <- gsub ("Acc", "Accelerometer", names(TidyextractData))
names(TidyextractData) <- gsub ("Gyro", "Gyroscope", names(TidyextractData))
names(TidyextractData) <- gsub ("BodyBody", "Body", names(TidyextractData))
names(TidyextractData) <- gsub ("Mag", "Magnitude", names(TidyextractData))
names(TidyextractData) <- gsub ("^t", "Time", names(TidyextractData))
names(TidyextractData)<- gsub("^f", "Frequency", names(TidyextractData))
names(TidyextractData) <- gsub("tBody", "TimeBody", names(TidyextractData))
names(TidyextractData) <- gsub("mean", ".Mean", names(TidyextractData),
        ignore.case = TRUE)
names(TidyextractData) <- gsub("std", "STD", names(TidyextractData), 
        ignore.case = TRUE)
names(TidyextractData) <- gsub("-freq()", "Frequency", names(TidyextractData), 
        ignore.case = TRUE)
names(TidyextractData) <- gsub("angle", "Angle", names(TidyextractData))
names(TidyextractData) <- gsub("gravity", "Gravity", names(TidyextractData))
```

## Now lets look at the new column names

```{r}
names(TidyextractData)
```

The now that our data is Tidy, our next task is to create a new data table that has the summary statistics for mean and standard deviation. I am going to create a new data table called "TidySummaryData" that has only the means of the mean and standard devation data. I will then write this file so that I can hand it in for the assignment. 

## Format data set to create a new data table with the average of each variable

```{r}
TidySummaryData <- TidyextractData %>%
        group_by(SubjectID, ActivityID) %>%
        summarise_all(funs(mean))
```


## Write the summary output to a text file             
```{r}
write.table(TidySummaryData, "TidySummaryData.txt", row.name = FALSE)

```

        

